---
title: "Soil vs Feces Microbial Diversity"
author: "Quintana_Theresa"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(42)
```

> **Data source:** Caporaso, J.G., et al. (2011). *Global patterns of 16S rRNA
> diversity at a depth of millions of sequences per sample.* PNAS 108:
> 4516–4522. Distributed as the `GlobalPatterns` example dataset in the
> `phyloseq` R package.


## Claim

Soil samples have higher microbial diversity (Shannon index) than human
feces samples, using the `GlobalPatterns` 16S dataset described above.


```{r load}
#load libraries
library(phyloseq) #organize microbiome data
library(vegan) #ecology math

data(GlobalPatterns)                                    # loads the built-in GlobalPatterns dataset into R session
otu <- as(otu_table(GlobalPatterns), "matrix")           # pulls out the OTU count table (taxa x samples) as a plain matrix
samp <- as(sample_data(GlobalPatterns), "data.frame")    # pulls out the sample metadata (e.g. SampleType) as a data frame
```

## Main analysis

```{r main}
shannon <- vegan::diversity(t(otu), index = "shannon")
df <- data.frame(sample = names(shannon), shannon = as.numeric(shannon),
                  type = samp[names(shannon), "SampleType"])

soil  <- df$shannon[df$type == "Soil"]
feces <- df$shannon[df$type == "Feces"]
observed_diff <- mean(soil) - mean(feces)

cat("Soil mean:", mean(soil), " Feces mean:", mean(feces), "\n")
wilcox.test(soil, feces)
```

## Control 1: Known answer

A perfectly even community of 4 taxa has Shannon = ln(4), by definition —
true regardless of this pipeline.

```{r known-answer}
cat("Expected: ln(4) =", log(4), "\n")
cat("Pipeline gives: ", vegan::diversity(rep(10, 4)), "\n")
```

## Control 2: Invariant

Shannon diversity must be non-negative for every sample, no matter what the
data say.

```{r invariant}
all(shannon >= 0)
```

## Control 3: Negative control

**Destroyed:** which label (Soil/Feces) is attached to which diversity
value. **Preserved:** the real diversity values themselves. Shuffle labels
1000 times, build a null distribution.

```{r negative-control}
sf <- df[df$type %in% c("Soil", "Feces"), ]
null_diffs <- replicate(1000, {
  shuffled <- sample(sf$type)
  mean(sf$shannon[shuffled == "Soil"]) - mean(sf$shannon[shuffled == "Feces"])
})
cat("Null mean:", mean(null_diffs), "(should be ~0)\n")
cat("Empirical p-value:", mean(abs(null_diffs) >= abs(observed_diff)), "\n")
hist(null_diffs, main = "Null distribution (labels scrambled)")
abline(v = observed_diff, col = "red", lwd = 2)
```

## Control 4: Positive control

**Planted:** take one real feces sample, truncate it to its top 20 taxa
(deliberately less diverse), and confirm the test finds that difference
against the untouched full community.

```{r positive-control}
base <- sort(otu[otu[, "M31Fcsw"] > 0, "M31Fcsw"], decreasing = TRUE)
p_full <- base / sum(base)
p_trunc <- rep(0, length(base)); p_trunc[1:20] <- base[1:20]; p_trunc <- p_trunc / sum(p_trunc)

A <- apply(rmultinom(5, 5000, p_full),  2, vegan::diversity)  # untouched
B <- apply(rmultinom(5, 5000, p_trunc), 2, vegan::diversity)  # planted low-diversity

cat("Full mean:", mean(A), " Truncated mean:", mean(B), "\n")
wilcox.test(A, B)
```

## Control 5: Redundancy

Same quantity, two ways: `vegan::diversity()` vs. a hand-written formula.
They share the same input data and the same math definition — but not the
same code.

```{r redundancy}
manual_shannon <- function(x) { p <- x[x > 0] / sum(x); -sum(p * log(p)) }
cat("vegan:  ", vegan::diversity(otu[, "CL3"]), "\n")
cat("manual: ", manual_shannon(otu[, "CL3"]), "\n")
```

## Control 6: Order of magnitude

Written before running anything above: soil diversity should be clearly higher than feces diversity, since soil habitats are generally far more microbially diverse than host-filtered gut communities — and both results (6.62 vs. 3.67) came out that way.

## Control 7: Determinism

```{r determinism}
r1 <- vegan::diversity(t(otu[, c("CL3", "M31Fcsw")]))
r2 <- vegan::diversity(t(otu[, c("CL3", "M31Fcsw")]))
identical(r1, r2)
```

*(Run the whole `.Rmd` twice / on another machine to confirm fully —
everything here is seeded.)*

## Environment

```{r env}
sessionInfo()
```